---
title: "p8105_hw1_yz4993"
author: "Yixin Zheng"
date: "2024-09-22"
output: github_document
---

# Problem 1

```{r 1.1 setup, include=FALSE}
#setting global option to show code for knitr
knitr::opts_chunk$set(echo = TRUE)
#load package for creating plots
library(ggplot2) 
#load package for penguins data
library(palmerpenguins) 
#loead package for creating dataframe
library(tidyverse)
```

```{r 1.2 examine dataset, include=FALSE}
#load penguins data set
data("penguins", package = "palmerpenguins")
#examine the variables
colnames(penguins) 
```

The `penguins` dataset contains `r nrow(penguins)` rows and `r ncol(penguins)` columns. The dataset contains the following variables:

-   `r colnames(penguins)[1]` is a factor denoting the penguins' species: `r paste(unique(penguins$species), collapse = ", ")`.

-   `r colnames(penguins)[2]` is a factor denoting the islands where penguins were found: `r paste(unique(penguins$island), collapse = ", ")`.

-   `r colnames(penguins)[3]` is a number denoting the bill length of penguin in millimeters, with a range from `r min(penguins$bill_length_mm, na.rm = TRUE)` mm to `r max(penguins$bill_length_mm, na.rm = TRUE)` mm.

-   `r colnames(penguins)[4]` is a number denoting the bill depth of penguin in millimeters, with a range from `r min(penguins$bill_depth_mm, na.rm = TRUE)` mm to `r max(penguins$bill_depth_mm, na.rm = TRUE)` mm.

-   `r colnames(penguins)[5]` is an integer denoting the flipper length of penguin in millimeters, with a range from `r min(penguins$flipper_length_mm, na.rm = TRUE)` mm to `r max(penguins$flipper_length_mm, na.rm = TRUE)` mm.

-   `r colnames(penguins)[6]` is an integer denoting the body mass of penguin in grams, with a range from `r min(penguins$body_mass_g, na.rm = TRUE)` g to `r max(penguins$body_mass_g, na.rm = TRUE)` g.

-   `r colnames(penguins)[7]` is a factor denoting the penguins sex: `r paste(unique(penguins$sex), collapse = ", ")`.

-   `r colnames(penguins)[8]` is an integer denoting the study year: `r paste(unique(penguins$year), collapse = ", ")`.

The mean flipper length is `r mean(penguins$flipper_length_mm, na.rm = TRUE)` mm.

```{r 1.3 scatter plot, warning=FALSE, message=FALSE}
#Suppressed warning and message outputs related to missing values and image saving (by changing options of code chunk)
#create a scatter plot with bill_length_mm on the x-axis and flipper_length_mm on the y-axis, color points based on species
ggplot(penguins, aes(x = bill_length_mm, y = flipper_length_mm, color = species)) +
#add points to plot
  geom_point() + 
#add labels and title and legend to the plot
  labs(
    title = "Scatterplot of Flipper Length vs Bill Length",
    x = "Bill Length (mm)",
    y = "Flipper Length (mm)",
    color = "Penguin Species"
  )

#save the scatterplot as a png file to the project directory
ggsave("penguin_scatterplot.png") 
```

# Problem 2
```{r 2.1 creating dataframe}
## set seed to ensure reproducibility
set.seed(1234)

## define data frame
df1 <- tibble(
  #a random sample of size 10 from a standard Normal distribution
  random_sample = rnorm(10),
  #a logical vector indicating whether elements of the sample are greater than 0
  positive = rnorm(10) > 0,
  #a character vector of length 10
  chr10 = letters[1:10],
  #a factor vector of length 10, with 3 factor levels (X,Y,Z) with replacement
  fac10_3 = factor(sample(c("X", "Y", "Z"), 10, replace = TRUE))
)

#view the dataframe
df1
```

```{r 2.2 calculate means first attempt}
#attempt to calculate the mean of each variable by pulling the variable out of the dataframe
mean_rand <- df1 %>% pull(random_sample) %>% mean()
mean_logi <- df1 %>% pull(positive) %>% mean()
mean_char <- df1 %>% pull(chr10) %>% mean()       
mean_fact <- df1 %>% pull(fac10_3) %>% mean()
```
We could take the mean for the random sample and logical vector when we try to take the mean directly. However, the warning message and the NA show that we could not take means of the character and the factor variable, therefore, we will need to try converting the variable type to numeric.

```{r 2.3 calculate means second attempt, results='hide'}
#attempt to calculate the mean by converting type
numeric_logi <- as.numeric(df1$positive) 
numeric_char <- as.numeric(df1$chr10)              
numeric_fact <- as.numeric(df1$fac10_3)
```

From the warning message we could know that the `as.numeric` function will work for the logical and factor vector by converting the logical vector from True/False to 1/0, and the factor by its integer codes. However, this function will not work for the character variable as `NAs introduced by coercion`. This clarifies why we cannot calculate means for the character variable: since characters cannot be converted to numeric values, R returns `NA`. The mean function calculates the average of a numeric vector, but it cannot operate on `NA` values, making it unsuitable for character variables.